### 2010-10-22 Thesis/Chapter4.tex
- Update 19 October -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@186 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 22 October -
Almost done...



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@188 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter{Modelling the deformation of solid objects in real-time}
\label{chap4}
\begin{shortAbstract}
Modelling the deformation of solid organs is a crucial problem in medical simulation. If the finite element method is classicaly used in mechanical engineering for its ability to solve the equations of continuum mechanics fairly accurately, its development in medical simulation was more progressive. The main reason for this is that the finite element method is computationally demanding and not easily compatible with time constraints required in interactive simulation. Therefore, other methods were developed to allow faster computations, at the cost of larger approximations of course. Over time, along with the increase in computational power, methods became more elaborate and less approximative. These different approaches may be grouped into three main categories: (1) geometrically based techniques, (2) approaches physically motivated and (3) methods actually based on the equations of continuum mechanics. In this chapter, we will review the most classic methods applied to modelling of solid objects with a particular emphasis on medical simulation. 
\end{shortAbstract}

\section{Introduction: the problematic}

@@

\ON
The challenges faced by the field of medical simulation were reviewed in the introductory chapter (see chapter \ref{chap1}). Hence, we know that at the very least, the deformation of solid organs must appear physically realist in medical simulators. In addition, this objective has to be achieved within heavy time constraints. Indeed, although each step of a simulation alone (deformation of organs, collision detection, contacts processing or visualisation) is computationally demanding, the whole simulation process must be computed in real-time or close to real-time. Of particular interest to us is the accurate modelling of the deformation of anatomical structures. The best mathematical model for predicting the deformation of a continuous structure is provided by continuum mechanics. If the finite element method is classicaly used in mechanical engineering for its ability to solve the equations of continuum mechanics fairly accurately, its development in medical simulation was more progressive. The main reason for this is that the finite element method is computationally very demanding and not easily compatible with the time constraints required in interactive simulation. Very often simplifications have to be made. Since medical simulation may find applications in various areas (such as medical training, patient-specific planning or per-operative guidance), the requirements in terms of accuracy as well as time constraints may be different. Consequently, a range of models has been developed over time to suit all usages.

For instance, the correctness of deformation is crucial for patient-specific planning and per-operative guidance. Let us give an example where simulation can lead the surgeon towards its goal during surgery: the removal of a brain tumor. Prior surgery, images of the patient's brain were acquired and the tumor was localised. To access the brain during surgery, the surgeon must first drill a hole into the patient's skull. However, drilling a hole releases intra-cranial pressure and this change in internal pressure deforms the brain. \OFF This phenomenon is called \emph{brain shift}. The overall deformation is quite complex and the displacements depend on local elasticity of the brain, the size and the location of the tumor. \ON Removing the tumor relies \OFF heavily on knowing the spatial relation between the patient's brain and the images acquired prior to surgery. \ON And because the brain deforms during surgery, the location and the shape of the tumor will no longer match those of the images acquired. \OFF Therefore, a large brain shift, if not corrected, will result in inaccuracies in the surgical procedure and has the potential to cause damage to normal tissue. A solution is to accurately model the mechanical response of the patient's brain to predict the actual deformation and hence keep the knowledge of the tumor's location. \ON If the accuracy is obviously of prime importance in this situation, computation of the deformation does not necessarily need to be carried out in real-time, a reasonable delay is acceptable (a couple of minutes).
\OFF

@@

Conversely, when medical simulation is applied to training, such a precision in terms of \ON deformation is not always \OFF required. \ON A visually realistic deformation is often sufficient for an inexperienced physician to learn the appropriate gestures required for a given medical procedure. However, in some cases, the feel of touch can be very important to master the procedure. \OFF For instance, during a colonoscopy procedure, force feedbacks give precious information about what type of loop is forming \ON during the colonoscope's insertion \OFF and the physician can then react accordingly to prevent loop formations. Consequently, a good colonoscopy simulator is required to provide accurate force feedbacks. If the contact forces were not reproduced to the operator, he would never be able to learn how to detect loop forming. Obviously, the computation of a realistic force to be returned demands an appropriate mechanical modelling of all structures. In constrast of per-operative guidance, medical training demands real-time computations of the deformations.

@@

Coquillard later developed the idea of Extented free-form deformation \citeyearpar{Coquillart90} and Animated free-form technique \citeyearpar{Coquillart91}. The idea of free-form deformation with lattices of arbitrary topology was introduced by \cite{MacCracken96}. Finally, \cite{Schein04} presented a technique to facilitate the incorporation of discontinuities in a model while deforming it.

	\subsection{Shape matching}
One field particulary interested in modelling deformable objects in a very efficient way is the game industry. Of course, for games we are more attracted by computational efficiency and extreme stability features than accuracy to physics laws. Most of the time, we can tolerate a degradation of realism as long as the result looks realistic. \cite{Muller05} developed a technique with this idea in mind called \emph{shape matching}. We start with a set of particules with masses $ m_i $ in an initial configuration where we denote by $ \mathbf{x}_i^0 $ the initial positions of the particules. No connectivity information is required. The particles are simulated as a simple particle system without particle-particle interactions, but including response to collisions with the environment and including external forces such as gravity. The positions in the deformed configuration are noted $ \mathbf{x}_i $. The method consists of finding a set of points $ \mathbf{g}_i $ which minimises the difference between the two sets $ \mathbf{x}_i^0 $ and $ \mathbf{x}_i $. The first step is to find the rotation matrix $ \mathbf{R} $ and the translation vectors $ \mathbf{t} $ and $ \mathbf{t}_0 $ such that
\begin{equation}
\sum_i m_i (\mathbf{R}(\mathbf{x}_i^0 - \mathbf{t}_0) + \mathbf{t} - \mathbf{x}_i)^2
\end{equation}
is minimal. The optimal translation vectors $ \mathbf{t} $ and $ \mathbf{t}_0 $ turn out to be the centre of mass of the initial shape and actual shape, that we will note $ \mathbf{x}_{c}^0 $ and $ \mathbf{x}_c $, respectively. The optimal rotation matrix is found by first finding the optimal linear transformation $ \mathbf{A} $. The optimal rotation is eventually obtained through the rotational part of $ \mathbf{A} $ after a polar decomposition. Finally, the goal positions $ \mathbf{g}_i $ can be computed as follows:
\begin{equation}
\label{chap4:gi}
\mathbf{g}_i = \mathbf{R}(\mathbf{x}_i^0 - \mathbf{x}_{c}^0) + \mathbf{x}_{c}.
\end{equation}
Once the goal positions are known, they are used to integrate the positions of the particules:
\begin{equation}  
	\begin{cases} 
		\mathbf{v}_i(t+h) = \mathbf{v}_i(t) + \alpha \dfrac{\mathbf{g}_i(t) - \mathbf{x}_i(t)}{h}  + h \dfrac{f_{\text{ext}}(t)}{m_i} \\\\
		\mathbf{x}_i(t+h) = \mathbf{x}_i(t) + h \mathbf{v}_i(t+h)
	\end{cases}
\end{equation}
where $ \alpha $ is a parameter between $ 0 $ and $ 1 $ which simulates stiffness. Because the method in this basic form is not suitable for objects undergoing large deformations, it is then extended to linear and quadratic deformations by replacing $ \mathbf{R} $ in \eqref{chap4:gi} with combinations of the form $ \beta \mathbf{A} + (1-\beta) \mathbf{R}$ where $ \beta $ is an additional parameter. Linear transformation can represent shear and stretch while quadratic transformation add twist and bending. Of course, this technique is not physically realistic. But this is beyond the point. This method allows the authors to simulate hundreds of deformable objects at an interactive rate with an unconditional stability (see \fig{chap4:fig-shoes}). Another illustration of this stability is shown \fig{chap4:fig-duck}. 
%
\begin{figure}[ht]
\begin{center}
\includegraphics[width=13.9cm]{chapter4/shoes.pdf}
\end{center}
\caption[Shape matching technique]{Shape matching allows to simulate hundreds of deformable objects in real-time. Image courtesy of \cite{Muller05}.}
\label{chap4:fig-shoes}
\end{figure}

@@

Later, \cite{Rivers07} extented the technique to regular lattices with embedded geometry. From a surface mesh to be deformed, they start by voxelising the model to construct a lattice of cubic cells containing the mesh. The embedded mesh is then deformed using trilinear interpolation of lattice vertex positions. A particule is placed at each vertex of the lattice and associated with a shape matching region. Their technique of \emph{fast lattice shape matching} allows them to compute per-region transformations in an efficient manner (see \fig{chap4:fig-penguins}).	
%
\begin{figure}[ht]
\centering 
\subfloat[ ]{\includegraphics[width=6cm]{chapter4/penguins1.pdf}}
\hfill
\subfloat[ ]{\includegraphics[width=6cm]{chapter4/penguins2.pdf}}
\caption[Penguins animation with fast lattice shape matching]{(a) $150$ cartoon penguins deforming dynamically using fast lattice shape matching (b). Deformed lattices consisting of $150$ particles per penguin ($22\,500$ particles). The penguins can be deformed at $25$ frames per second on a Pentium4 $3.4 \,$GHz. Image courtesy of \cite{Rivers07}. }
\label{chap4:fig-penguins}
\end{figure}

@@

This method was extented by \cite{Schill98} to integrate inhomogeneous and anisotropic behaviour. This extension of the chain mail algorithm allows the authors to simulate the vitrous humor in the eye during a vitrectomy procedure. However, the main drawback of this algorithm is that it does not model volume preservation, an important characteristic of many tissues in the human body.

@@

\cite{James02} went even further by using graphics hardware to reduce CPU costs to a minimum. In particular, they use modal analysis to model human skin and secondary tissues in a laparascopic surgical simulation to allow the CPU to focus on more complex tissue models and user contact interactions (see \fig{chap4:fig-modalAnalysisLaparo}). The technique is fairly efficient as it allows precomputations of the vibration modes. 
%
\begin{figure}[ht]
\centering 
\subfloat[ ]{\includegraphics[height=6cm]{chapter4/modalAnalysisLaparo1.pdf}}
\hspace{0.2cm}
\subfloat[ ]{\includegraphics[height=6cm]{chapter4/modalAnalysisLaparo2.pdf}}
\caption[Applications of modal analysis]{(a) A jumping motion that leads to significant thigh and belly vibrations of human skin when rendered with modal analysis. (b) The same hardware accelerated technique is applied to secondary tissues in a laparoscopic simulation to ease the CPU of computations. Image courtesy of \cite{James02}.} 
\label{chap4:fig-modalAnalysisLaparo}
\end{figure}

Although modal analysis significantly accelerates the simulation, it generates noticeable artifacts when applied to large deformations due to the step of linearisation. It can produce unnatural results with deformations of large amplitude. \cite{Choi05a} proposed to overcome this limitation by taking the rotational component of the deformation into account at each node. \cite{Barbic05} make use of modal analysis to build a deformation basis from modal derivatives. It allows them to simulate large deformations of solid objects with non-linear material.

	\subsection{Mass-spring model}	\label{chap4:massSprings}
A classic technique for modelling a deformable object is the mass-spring system. In this method, the geometry is described by a network of masses connected together by springs. The force $ \mathbf{F} $ exerted onto each mass by the spring may be computed with:
\begin{equation}
\label{chap4:massSpring1}
\mathbf{F}_{\text{spring}} = - k \mathbf{x},
\end{equation}
where $ k $ is the spring stiffness constant, $ \mathbf{x} $ represents the stretch of the spring from its rest position. From our experience, we know that a mass attached to a single spring has a tendency to oscillate. If no appropriate measure is taken, a network of springs will a forciori oscillate as well. To prevent the system from oscillating, a damper is also added between each couple of masses. The force created by the damper acts to reduce the velocity and may be expressed as:
\begin{equation}
\label{chap4:massSpring2}
\mathbf{F}_{\text{damper}} = - d \mathbf{\dot x}
\end{equation}
where $ d $ is the damping constant and $ \mathbf{\dot x} $ the derivative of $ x $ with respect to time. The damper added between each couple of masses allows to model friction and reduce the amplitude of the oscillations. Applying Newton's second law $ \sum \mathbf{F} = m \mathbf{\ddot x} $ yields the standard differential equation for a mass-spring system:
\begin{equation}
\label{chap4:massSpring3}
m \mathbf{\ddot x} = - d \mathbf{\dot x} - k \mathbf{x}.
\end{equation}
For complex systems with more than one spring attached to each mass, the right-hand side of the equation \eqref{chap4:massSpring3} becomes more complex as it needs to be repeated and adjusted for each spring. The easiest way to solve a complex mass-spring system iteratively, is to first calculate the total force currently acting on each mass by summing all forces currently exerted by all springs attached to this mass (using \eqref{chap4:massSpring1} and \eqref{chap4:massSpring2}). Dividing the total force by the mass gives the current acceleration of each mass (by \eqref{chap4:massSpring3}). This equation must be numerically solved for each mass, either by explicit or implicit methods. A discussion of integration methods for mass-spring models can be found in \cite{Shinya05}.

@@

Because they do not derive from the equations of continuum mechanics, mass-spring models have a limited capacity to model the various aspects of a material like anisotropy, viscoelasticity etc. Even worse, according to \cite{Bourguignon00}, if all springs are set to the same stiffness, the mesh geometry may generate undesired anisotropy as shown in \fig{chap4:fig-undesiredAnisotropy}. However, some works tried to take advantage of this feature by giving directions of interest to their model by specifically designing the mesh to align springs on specific directions. For instance, it was used by \cite{Miller88} to animate the motion of snakes and worms. It was also used by \cite{Ng-Thow-Hing97} in their muscle model where some of the springs were aligned with the muscle fibers and more recently by \cite{Magjarevic09} for modelling the mechanical behavior of the myocardial tissue. More details on controlling anisotropy in mass-spring systems may be found in \cite{Bourguignon00}. 
%
\begin{figure}[ht]
\centering 
\subfloat[ ]{\includegraphics[width=4cm]{chapter4/undesiredAnisotropy1.pdf}}
\hspace{0.2cm}
\subfloat[ ]{\includegraphics[width=4cm]{chapter4/undesiredAnisotropy2.pdf}}
\caption[Undesired anisotropy with mass-spring models]{The two meshes are undergoing a downward pull at their bottom. While the tetrahedral mass-spring system (a) shows signs of anisotropy, the hexahedral mesh (b) with springs aligned along the gravity does not. Image courtesy of \cite{Bourguignon00}. } 
\label{chap4:fig-undesiredAnisotropy}
\end{figure}

A mass-spring model may also be enhanced to feature viscoelasticity. Indeed, \cite{Stiles72} proposed a viscoelastic mass-spring system to test hypotheses on the mechanical response of muscles. Much more recently, \cite{Tamura05} introduced a method to simulate viscoelastic material using a mass-spring system. They use a large number of particules to create a randomly connected mesh to mimic the structure of polymeric materials and hence their viscoelascity characteristics. \cite{Basafa10} went even further and extended a mass-spring system to simulate non-linear viscoelastic deformations of soft tissue for laparoscopic surgery. They tune their parameters by a simple optimisation procedure to fit the mechanical response obtained on a set of experimental data.

@@

Taking advantage of the GPU to reach a substantial speed-up is usually not straightforward. One must understand the limitations inherent in its design and devise algorithms accordingly. With this idea in mind, \cite{Sorensen06} reviewed the most important concepts and data structures required to realise two popular deformable models on the GPU: the spring-mass model and the finite element model. \ON In 2008, \citeauthor{Rasmusson08} investigated performance and optimisation strategies between the two GPU languages OpenGL and CUDA for their mass-spring-damper model. More recently, \cite{Leon10} presented a GPU implementation of a mass-spring model with a nearly $80\times$ speedup over serial execution and $20\times$ speedup over CPU-based parallel execution. \OFF

@@

In 1996, \citeauthor{Bro-Nielsen96} developed a real-time finite element method formulation by introducing three improvements. The first one is to compress the stiffness matrix of the volumetric system into a matrix with the same complexity as a surface model of the same object. This technique is called \emph{condensation}. The deformation is solved only at surface points while still taking the volumetric nature of the object into account. The second improvement concerns the fact that they explicitly precompute the inverse of the stiffness matrix and use matrix vector multiplication with this matrix to achieve a low calculation time. Finally, they exploit the sparse structure of the force vector. They showed the efficiency of their method by modelling a leg with 700 nodes at 20 frames per second.

@@

\cite{Cotin99} took advantage of the superposition principle which applies to linear problems. Their idea was to approximate the deformation of a solid object with a linear combination of pre-computed states. In addition, they enhanced the formulation by adding a non-linear radial component. This method was applied to compute the deformation of liver in a hepatic surgery simulator. Their approach substantially improved the computational time for a finite element model as it allowed them to compute the positions and the forces of a $1\,500$ node liver at $300\,$Hz. An important drawback emphasised by the authors themselves is the impossibility to simulate tissue cutting. Indeed, the large precomputations associated with this method are dependent of the geometry of the mesh and therefore cannot be updated in real-time.

>>>>

\chapter[Modelling the deformation of solid objects in real-time][Modelling solid objects in real-time]{Modelling the deformation of solid objects in real-time}
\label{chap4}
\begin{shortAbstract}
Modelling the deformation of solid organs is a crucial problem in medical simulation. If the finite element method is classicaly used in mechanical engineering for its ability to solve the equations of continuum mechanics fairly accurately, its development in medical simulation was more progressive. The main reason for this is that the finite element method is computationally demanding and not easily compatible with time constraints required in interactive simulation. Therefore, other methods were developed to allow faster computations, at the cost of larger approximations of course. Over time, along with the increase in computational power, methods became more elaborate and less approximative. These different approaches may be grouped into three main categories: (1) geometrically based techniques, (2) approaches physically motivated and (3) methods actually based on the equations of continuum mechanics. In this chapter, we will review the most classic methods applied to modelling of solid objects with a particular emphasis on medical simulation. 
\end{shortAbstract}

\section{Introduction: the problem}

@@

The challenges faced by the field of medical simulation were reviewed in the introductory chapter (see chapter \ref{chap1}). Hence, we know that at the very least, the deformation of solid organs must appear physically realist in medical simulators. In addition, this objective has to be achieved within heavy time constraints. Indeed, although each step of a simulation alone (deformation of organs, collision detection, contacts processing or visualisation) is computationally demanding, the whole simulation process must be computed in real-time or close to real-time. Of particular interest to us is the accurate modelling of the deformation of anatomical structures. The best mathematical model for predicting the deformation of a continuous structure is provided by continuum mechanics. If the finite element method is classically used in mechanical engineering for its ability to solve the equations of continuum mechanics fairly accurately, its development in medical simulation was more progressive. The main reason for this is that the finite element method is computationally very demanding and not easily compatible with the time constraints required in interactive simulation. Very often simplifications have to be made. Since medical simulation may find applications in various areas (such as medical training, patient-specific planning or per-operative guidance), the requirements in terms of accuracy as well as time constraints may be different. Consequently, a range of models has been developed over time to suit all usages.

For instance, the correctness of deformation is crucial for patient-specific planning and per-operative guidance. Let us give an example where simulation can lead the surgeon towards its goal during surgery: the removal of a brain tumour. Prior surgery, images of the patient's brain were acquired and the tumour was localised. To access the brain during surgery, the surgeon must first drill a hole into the patient's skull. However, drilling a hole releases intra-cranial pressure and this change in internal pressure deforms the brain. This phenomenon is called \emph{brain shift}. The overall deformation is quite complex and the displacements depend on local elasticity of the brain, the size and the location of the tumour. Removing the tumour relies heavily on knowing the spatial relation between the patient's brain and the images acquired prior to surgery. And because the brain deforms during surgery, the location and the shape of the tumour will no longer match those of the images acquired. Therefore, a large brain shift, if not corrected, will result in inaccuracies in the surgical procedure and has the potential to cause damage to normal tissue. A solution is to accurately model the mechanical response of the patient's brain to predict the actual deformation and hence keep the knowledge of the tumour's location. If the accuracy is obviously of prime importance in this situation, computation of the deformation does not necessarily need to be carried out in real-time, a reasonable delay is acceptable (a couple of minutes).

@@

Conversely, when medical simulation is applied to training, such a precision in terms of deformation is not always required. A visually realistic deformation is often sufficient for an inexperienced physician to learn the appropriate gestures required for a given medical procedure. However, in some cases, the feel of touch can be very important to master the procedure. For instance, during a colonoscopy procedure, force feedbacks give precious information about what type of loop is forming during the colonoscope's insertion and the physician can then react accordingly to prevent loop formations. Consequently, a good colonoscopy simulator is required to provide accurate force feedbacks. If the contact forces were not reproduced to the operator, he would never be able to learn how to detect loop forming. Obviously, the computation of a realistic force to be returned demands an appropriate mechanical modelling of all structures. In constrast of per-operative guidance, medical training demands real-time computations of the deformations.

@@

Coquillard later developed the idea of extended free-form deformation \citep{Coquillart90} and animated free-form technique \citep{Coquillart91}. The idea of free-form deformation with lattices of arbitrary topology was introduced by \cite{MacCracken96}. Finally, \cite{Schein04} presented a technique to facilitate the incorporation of discontinuities in a model while deforming it.

	\subsection{Shape matching}
One field particularly interested in modelling deformable objects in a very efficient way is the game industry. Of course, for games we are more attracted by computational efficiency and extreme stability features than accuracy to physics laws. Most of the time, we can tolerate a degradation of realism as long as the result looks realistic. \cite{Muller05} developed a technique with this idea in mind called \emph{shape matching}. We start with a set of particles with masses $ m_i $ in an initial configuration where we denote by $ \mathbf{x}_i^0 $ the initial positions of the particles. No connectivity information is required. The particles are simulated as a simple particle system without particle-particle interactions, but including response to collisions with the environment and including external forces such as gravity. The positions in the deformed configuration are noted $ \mathbf{x}_i $. The method consists of finding a set of points $ \mathbf{g}_i $ which minimises the difference between the two sets $ \mathbf{x}_i^0 $ and $ \mathbf{x}_i $. The first step is to find the rotation matrix $ \mathbf{R} $ and the translation vectors $ \mathbf{t} $ and $ \mathbf{t}_0 $ such that
\begin{equation}
\sum_i m_i (\mathbf{R}(\mathbf{x}_i^0 - \mathbf{t}_0) + \mathbf{t} - \mathbf{x}_i)^2
\end{equation}
is minimal. The optimal translation vectors $ \mathbf{t} $ and $ \mathbf{t}_0 $ turn out to be the centre of mass of the initial shape and actual shape, that we will note $ \mathbf{x}_{c}^0 $ and $ \mathbf{x}_c $, respectively. The optimal rotation matrix is found by first finding the optimal linear transformation $ \mathbf{A} $. The optimal rotation is eventually obtained through the rotational part of $ \mathbf{A} $ after a polar decomposition. Finally, the goal positions $ \mathbf{g}_i $ can be computed as follows:
\begin{equation}
\label{chap4:gi}
\mathbf{g}_i = \mathbf{R}(\mathbf{x}_i^0 - \mathbf{x}_{c}^0) + \mathbf{x}_{c}.
\end{equation}
Once the goal positions are known, they are used to integrate the positions of the particules:
\begin{equation}  
	\begin{cases} 
		\mathbf{v}_i(t+h) = \mathbf{v}_i(t) + \alpha \dfrac{\mathbf{g}_i(t) - \mathbf{x}_i(t)}{h}  + h \dfrac{f_{\text{ext}}(t)}{m_i} \\\\
		\mathbf{x}_i(t+h) = \mathbf{x}_i(t) + h \mathbf{v}_i(t+h)
	\end{cases}
\end{equation}
where $ \alpha $ is a parameter between $ 0 $ and $ 1 $ which simulates stiffness. Because the method in this basic form is not suitable for objects undergoing large deformations, it is then extended to linear and quadratic deformations by replacing $ \mathbf{R} $ in \eqref{chap4:gi} with combinations of the form $ \beta \mathbf{A} + (1-\beta) \mathbf{R}$ where $ \beta $ is an additional parameter. Linear transformation can represent shear and stretch while quadratic transformation add twist and bending. Of course, this technique is not physically realistic. But this is beyond the point. This method allows the authors to simulate hundreds of deformable objects at an interactive rate with an unconditional stability (see \fig{chap4:fig-shoes}). Another illustration of this stability is shown \fig{chap4:fig-duck}. 
%
\begin{figure}[ht]
\begin{center}
\includegraphics[width=13.9cm]{chapter4/shoes.pdf}
\end{center}
\caption[Shape matching technique]{Shape matching allows to simulate hundreds of deformable objects in real-time. Image courtesy of \cite{Muller05}.}
\label{chap4:fig-shoes}
\end{figure}

@@

Later, \cite{Rivers07} extended the technique to regular lattices with embedded geometry. From a surface mesh to be deformed, they start by voxelising the model to construct a lattice of cubic cells containing the mesh. The embedded mesh is then deformed using trilinear interpolation of lattice vertex positions. A particle is placed at each vertex of the lattice and associated with a shape matching region. Their technique of \emph{fast lattice shape matching} allows them to compute per-region transformations in an efficient manner (see \fig{chap4:fig-penguins}).	
%
\begin{figure}[ht]
\centering 
\subfloat[ ]{\includegraphics[width=6cm]{chapter4/penguins1.pdf}}
\hfill
\subfloat[ ]{\includegraphics[width=6cm]{chapter4/penguins2.pdf}}
\caption[Penguins animation with fast lattice shape matching]{(a) $150$ cartoon penguins deforming dynamically using fast lattice shape matching (b). Deformed lattices consisting of $150$ particles per penguin ($22\,500$ particles). The penguins can be deformed at $25$ frames per second on a Pentium4 $3.4 \,$GHz. Image courtesy of \cite{Rivers07}. }
\label{chap4:fig-penguins}
\end{figure}

@@

This method was extended by \cite{Schill98} to integrate inhomogeneous and anisotropic behaviour. This extension of the chain mail algorithm allows the authors to simulate the vitreous humour in the eye during a vitrectomy procedure. However, the main drawback of this algorithm is that it does not model volume preservation, an important characteristic of many tissues in the human body.

@@

\cite{James02} went even further by using graphics hardware to reduce CPU costs to a minimum. In particular, they use modal analysis to model human skin and secondary tissues in a laparascopic surgical simulation to allow the CPU to focus on more complex tissue models and user contact interactions (see \fig{chap4:fig-modalAnalysisLaparo}). The technique is fairly efficient as it allows pre-computations of the vibration modes. 
%
\begin{figure}[ht]
\centering 
\subfloat[ ]{\includegraphics[height=6cm]{chapter4/modalAnalysisLaparo1.pdf}}
\hspace{0.2cm}
\subfloat[ ]{\includegraphics[height=6cm]{chapter4/modalAnalysisLaparo2.pdf}}
\caption[Applications of modal analysis]{(a) A jumping motion that leads to significant thigh and belly vibrations of human skin when rendered with modal analysis. (b) The same hardware accelerated technique is applied to secondary tissues in a laparoscopic simulation to ease the CPU of computations. Image courtesy of \cite{James02}.} 
\label{chap4:fig-modalAnalysisLaparo}
\end{figure}

Although modal analysis significantly accelerates the simulation, it generates noticeable artefacts when applied to large deformations due to the step of linearisation. It can produce unnatural results with deformations of large amplitude. \cite{Choi05a} proposed to overcome this limitation by taking the rotational component of the deformation into account at each node. \cite{Barbic05} make use of modal analysis to build a deformation basis from modal derivatives. It allows them to simulate large deformations of solid objects with non-linear material.

	\subsection{Mass-spring model}	\label{chap4:massSprings}
A classic technique for modelling a deformable object is the mass-spring system. In this method, the geometry is described by a network of masses connected together by springs. The force $ \mathbf{F} $ exerted onto each mass by the spring may be computed with:
\begin{equation}
\label{chap4:massSpring1}
\mathbf{F}_{\text{spring}} = - k \mathbf{x},
\end{equation}
where $ k $ is the spring stiffness constant, $ \mathbf{x} $ represents the stretch of the spring from its rest position. From our experience, we know that a mass attached to a single spring has a tendency to oscillate. If no appropriate measure is taken, a network of springs will a fortiori oscillate as well. To prevent the system from oscillating, a damper is also added between each couple of masses. The force created by the damper acts to reduce the velocity and may be expressed as:
\begin{equation}
\label{chap4:massSpring2}
\mathbf{F}_{\text{damper}} = - d \mathbf{\dot x}
\end{equation}
where $ d $ is the damping constant and $ \mathbf{\dot x} $ the derivative of $ x $ with respect to time. The damper added between each couple of masses allows to model friction and reduce the amplitude of the oscillations. Applying Newton's second law $ \sum \mathbf{F} = m \mathbf{\ddot x} $ yields the standard differential equation for a mass-spring system:
\begin{equation}
\label{chap4:massSpring3}
m \mathbf{\ddot x} = - d \mathbf{\dot x} - k \mathbf{x}.
\end{equation}
For complex systems with more than one spring attached to each mass, the right-hand side of the equation \eqref{chap4:massSpring3} becomes more complex as it needs to be repeated and adjusted for each spring. The easiest way to solve a complex mass-spring system iteratively, is to first calculate the total force currently acting on each mass by summing all forces currently exerted by all springs attached to this mass (using \eqref{chap4:massSpring1} and \eqref{chap4:massSpring2}). Dividing the total force by the mass gives the current acceleration of each mass (by \eqref{chap4:massSpring3}). This equation must be numerically solved for each mass, either by explicit or implicit methods. A discussion of integration methods for mass-spring models can be found in \cite{Shinya05}.

@@

Because they do not derive from the equations of continuum mechanics, mass-spring models have a limited capacity to model the various aspects of a material like anisotropy, viscoelasticity etc. Even worse, according to \cite{Bourguignon00}, if all springs are set to the same stiffness, the mesh geometry may generate undesired anisotropy as shown in \fig{chap4:fig-undesiredAnisotropy}. However, some works tried to take advantage of this feature by giving directions of interest to their model by specifically designing the mesh to align springs on specific directions. For instance, it was used by \cite{Miller88} to animate the motion of snakes and worms. It was also used by \cite{Ng-Thow-Hing97} in their muscle model where some of the springs were aligned with the muscle fibers and more recently by \cite{Magjarevic09} for modelling the mechanical behaviour of the myocardial tissue. More details on controlling anisotropy in mass-spring systems may be found in \cite{Bourguignon00}. 
%
\begin{figure}[ht]
\centering 
\subfloat[ ]{\includegraphics[width=4cm]{chapter4/undesiredAnisotropy1.pdf}}
\hspace{0.2cm}
\subfloat[ ]{\includegraphics[width=4cm]{chapter4/undesiredAnisotropy2.pdf}}
\caption[Undesired anisotropy with mass-spring models]{The two meshes are undergoing a downward pull at their bottom. While the tetrahedral mass-spring system (a) shows signs of anisotropy, the hexahedral mesh (b) with springs aligned along the gravity does not. Image courtesy of \cite{Bourguignon00}. } 
\label{chap4:fig-undesiredAnisotropy}
\end{figure}

A mass-spring model may also be enhanced to feature viscoelasticity. Indeed, \cite{Stiles72} proposed a viscoelastic mass-spring system to test hypotheses on the mechanical response of muscles. Much more recently, \cite{Tamura05} introduced a method to simulate viscoelastic material using a mass-spring system. They use a large number of particles to create a randomly connected mesh to mimic the structure of polymeric materials and hence their viscoelascity characteristics. \cite{Basafa10} went even further and extended a mass-spring system to simulate non-linear viscoelastic deformations of soft tissue for laparoscopic surgery. They tune their parameters by a simple optimisation procedure to fit the mechanical response obtained on a set of experimental data.

@@

Taking advantage of the GPU to reach a substantial speed-up is usually not straightforward. One must understand the limitations inherent in its design and devise algorithms accordingly. With this idea in mind, \cite{Sorensen06} reviewed the most important concepts and data structures required to realise two popular deformable models on the GPU: the spring-mass model and the finite element model. In 2008, \citeauthor{Rasmusson08} investigated performance and optimisation strategies between the two GPU languages OpenGL and CUDA for their mass-spring-damper model. More recently, \cite{Leon10} presented a GPU implementation of a mass-spring model with a nearly $80\times$ speedup over serial execution and $20\times$ speedup over CPU-based parallel execution.

@@

In 1996, \citeauthor{Bro-Nielsen96} developed a real-time finite element method formulation by introducing three improvements. The first one is to compress the stiffness matrix of the volumetric system into a matrix with the same complexity as a surface model of the same object. This technique is called \emph{condensation}. The deformation is solved only at surface points while still taking the volumetric nature of the object into account. The second improvement concerns the fact that they explicitly pre-compute the inverse of the stiffness matrix and use matrix vector multiplication with this matrix to achieve a low calculation time. Finally, they exploit the sparse structure of the force vector. They showed the efficiency of their method by modelling a leg with 700 nodes at 20 frames per second.

@@

\cite{Cotin99} took advantage of the superposition principle which applies to linear problems. Their idea was to approximate the deformation of a solid object with a linear combination of pre-computed states. In addition, they enhanced the formulation by adding a non-linear radial component. This method was applied to compute the deformation of liver in a hepatic surgery simulator (see \fig{chap4:fig-superposition}). Their approach substantially improved the computational time for a finite element model as it allowed them to compute the positions and the forces of a $1\,500$ node liver at $300\,$Hz. An important drawback emphasised by the authors themselves is the impossibility to simulate tissue cutting. Indeed, the large pre-computations associated with this method are dependent of the geometry of the mesh and therefore cannot be updated in real-time. 
%
\begin{figure}[ht]
\centering 
\includegraphics[width=13.5cm]{chapter4/superposition.pdf}
\caption[Hepatic surgery simulator]{Images of a hepatic surgery simulator. The liver model contains about $ 1\,500 $ nodes. The simulation was performed at $ 50\,$Hz for the visual feedback and $ 300\,$Hz for force feedback. Image courtesy of \cite{Cotin99}.}
\label{chap4:fig-superposition}
\end{figure}

\cite{Cotin00} proposed a hybrid model allowing real-time cutting. They combined a quasi-static precomputed model, which is very efficient but does not allow topology changes, with a tensor-mass model which requires more computation but authorises cutting. The key idea is to separate a same structure into two parts: one susceptible to be cut during the surgery procedure and one that will not be submitted to any topological change. While the former will be simulated with the tensor-mass model, the deformation of the latter will be computed using the precomputed model. Since both linearly elastic models follow the same physical law, the combination of these two models should behave exactly as a global, linearly elastic model. To achieve this goal, \citeauthor{Cotin00} describe how to impose the additional boundary conditions at the connection nodes for the global model to be consistent in terms of forces and displacements. The efficiency of this hybrid model was demonstrated by simulating an hepatectomy (removing of one of the eight anatomical segments of the liver). They reported the simulation (allowing deformation and cutting) of a mesh with more than $ 8\,000 $ tetrahedra in real-time.
% (see \fig{chap4:fig-hybridModel})
%\begin{figure}[ht]
%\centering 
%\subfloat[ ]{\includegraphics[height=4.5cm]{chapter4/hybrid1.pdf}}
%\hspace{0.2cm}
%\subfloat[ ]{\includegraphics[height=4.5cm]{chapter4/hybrid2.pdf}}
%\caption[Simulation of a hepatectomy using a hybrid model]{Simulation of a hepatectomy using a hybrid model. $ 18 $\% of the mesh was modelled with the tensor-mass model (green) and therefore could be cut out. The rest of the mesh was simulated with the precomputed finite element model. Image courtesy of \cite{Cotin00}.} 
%\label{chap4:fig-hybridModel}
%\end{figure}
